Finally Tagless, Partially Evaluated

### 2007-09-01 tagless/tagless-final.tex
removing section 4.3; all code in Sec 4 is OCaml

\section{A tagless partial evaluator}\label{PE}

Surprisingly, we can write a partial evaluator using the idea above,
namely to build object terms using ordinary functions rather than data
constructors.  We present this partial evaluator in a sequence of four
attempts. It uses no universal type and no tags
for object types.  We then discuss residualization and binding-time
analysis.  Our partial evaluator is a modular extension of the evaluator
in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{compiler}, in
that it uses the former to reduce static terms and the latter to build
dynamic terms.

@@

\ifshort\else
Our typed partial evaluator is online and polyvariant.  It reuses the
compiler~|C| and the evaluator~|R| by composing them.  This situation is
simpler than \citets{SperberThiemann:TwoForOne} composition of a partial
evaluator and a compiler, but the general ideas are similar.
\fi

>>>>

%------------------------------------------------------------------------
% This is the version for the APLAS paper. No Haskell
\ifshort
\section{A tagless partial evaluator}\label{PE}
Surprisingly, we can write a partial evaluator using the idea above,
namely to build object terms using ordinary functions rather than data
constructors.  We present this partial evaluator in a sequence of four
attempts. It uses no universal type and no tags
for object types.  We then discuss residualization and binding-time
analysis.  Our partial evaluator is a modular extension of the evaluator
in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{compiler}, in
that it uses the former to reduce static terms and the latter to build
dynamic terms.

Roughly, a partial evaluator interprets each object term to either
a static (present-stage) term (using\ifshort\else\ the evaluator\fi~|R|) or
a dynamic (future-stage) term (using\ifshort\else\ the
compiler\fi~|C|). To
distinguish between static and dynamic terms, one might define
|repr| in the first version of the partial evaluator |P0| as
\texttt{type ('c,'dv) repr 
	= S0 of ('c,'dv) R.repr~}\Verb+|+\texttt{~E0 of ('c,'dv) C.repr}.
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R.add|) if and only 
if both operands are static; otherwise we extract the dynamic terms 
from the operands and add them using~|C.add|. We use |C.int| to
convert from the static term |('c,int) R.repr|, which is just |int|,
to the dynamic term.
Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(of type |('c,bool) C.repr|), the second a static term 
(of type |('c,'a) R.repr|) and the third a
dynamic term (of type |('c,'a) C.repr|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |'a -> ('c,'a) C.repr|, to send a value to the future stage
\citep{xi-guarded,WalidPOPL03}.

Our |Symantics| only includes separate lifting methods |bool| and
|int|, not a parametrically polymorphic lifting method, for good reason:
When compiling to a first-order target language such as machine code,
booleans, integers, and functions may well be represented differently.
Thus, compiling polymorphic lift requires intensional type
analysis.  To avoid needing polymorphic lift, we turn to
Asai's technique \citep{asai-binding-time,sumii-hybrid}:
build a dynamic term
alongside every static term.

We switch to the data type
\texttt{type ('c,'dv) repr = P1 of ('c,'dv) R.repr option * ('c,'dv) C.repr}
so that a partially evaluated term always contains a dynamic
component and sometimes contains a static component.  By 
distributivity, the two
alternative constructors of an |option| value, |Some| and |None|,
tag each partially evaluated term with a phase: either present or
future.  This tag is not an object type tag: all pattern matching below
is exhaustive.  Because the future-stage component is always present, we
can now define the polymorphic function
|let abstr1 (P1 (_,dyn)) = dyn| of type
\texttt{abstr1 : ('c,'dv) repr -> ('c,'dv) C.repr}
to extract it without requiring polymorphic lift into~|C|.  We then try
to define the interpreter |P1|---and get as far as the first-order
constructs of our object language, including |if_|.
\vspace*{-0.7em}
\begin{code3}
module P1 : Symantics = struct
  let int  (x:int)  = P1 (Some (R.int x), C.int x)
  let add e1 e2 = match (e1,e2) with
    | (P1 (Some n1,_),P1 (Some n2,_)) -> int (R.add n1 n2)
    | _ -> P1 (None,(C.add (abstr1 e1) (abstr1 e2)))
  let if_ = function
    | P1 (Some s,_) -> fun et ee -> if s then et () else ee ()
    | eb -> fun et ee -> P1 (None, C.if_ (abstr1 eb) 
                                   (fun () -> abstr1 (et ()))
                                   (fun () -> abstr1 (ee ())))
\end{code3}
\vspace*{-0.7em}
When we come to functions, however, we stumble.  According to our
definition of~|P1|, a partially evaluated object function, such as the
identity $\fun{x}x$ embedded in OCaml as |lam (fun x -> x)|\texttt{ :
}|('c,'a->'a) P1.repr|, consists of a dynamic part 
(type |('c,'a->'a) C.repr|) and
maybe a static part (type |('c,'a->'a) R.repr|).  The dynamic part is useful
when this function is passed to another function that is only
dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
useful when this function is applied to a static argument, as in
$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
evaluate the function, that is, compute as much as possible statically
when it is applied to a mix of static and dynamic inputs.  For example,
the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
not statically known.  The same static function, applied to
different static arguments, can give both static and dynamic results: we
want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
to~$x$.

To enable these simplifications, we delay binding-time analysis
for a static function until it is applied, that is, until |lam f|
appears as the argument of |app|.  To do so, we have to incorporate |f|
as it is into the |P1.repr| data structure: the representation
for a function type |'a->'b| should be either
\vspace*{-0.7em}
\begin{code}
S1 of ('c,'a) repr -> ('c,'b) repr | E1 of ('c,'a->'b) C.repr
P1 of (('c,'a) repr -> ('c,'b) repr) option * ('c,'a->'b) C.repr
\end{code}
\vspace*{-0.7em}
unlike |P1.repr| of |int| or |bool|.
That is, we need a nonparametric data type, something akin to
type-indexed functions and type-indexed types, which
\citet{oliveira-typecase} dub the \emph{typecase} design pattern.
Thus, typed partial evaluation, like typed CPS transformation,
inductively defines a map from source types to target types that
performs case distinction on the source type. In Haskell, typecase
can be equivalently implemented either with GADTs or 
type-class functional dependencies
\citep{oliveira-typecase}. The accompanying code shows both
approaches, neither of which is portable to OCaml. In addition,
the problem of nonexhaustive pattern\hyp matching reappears in
in the GADT approach because GHC 6.6.1 cannot see that a particular
type of a GADT value precludes certain constructors.
Recall, our
goal is to make this exhaustiveness \emph{syntactically} apparent.

Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
\texttt{type ('c,'dv) repr = P1 of repr\_pe ('c,'dv) R.repr option * ('c,'dv) C.repr}
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by
\vspace*{-0.8em}
\begin{code3}
repr_pe ('c,int) = ('c,int); repr_pe ('c,bool) = ('c,bool)
repr_pe ('c,'a->'b) = ('c,'a) repr -> ('c,'b) repr
\end{code3}
\vspace*{-0.8em}
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and portable solution thus emerges: we bake |repr_pe| 
into the signature |Symantics|. 
We recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that the |repr| type
constructor took two arguments |'c| and~|'dv|. We add an argument
|'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new signature.
% aplas: abbreviated figure
\begin{figure}
\begin{floatrule}
\ifshort
\begin{code2}
module type Symantics = sig type ('c,'sv,'dv) repr
  val int : int -> ('c,int,int) repr
  val lam : (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
            -> ('c,'x,'da -> 'db) repr
  val app : ('c,'x,'da -> 'db) repr
            -> (('c,'sa,'da) repr -> ('c,'sb,'db) repr as 'x)
  val fix : ('x -> 'x) -> (('c, ('c,'sa,'da) repr -> ('c,'sb,'db) repr,
                                'da -> 'db) repr as 'x)
  val add : ('c,int,int) repr -> ('c,int,int) repr -> ('c,int,int) repr
  val if_ : ('c,bool,bool) repr
            -> (unit->'x) -> (unit->'x) -> (('c,'sa,'da) repr as 'x) end
\end{code2}
\else
\begin{code2}
module type Symantics = sig
  type ('c,'sv,'dv) repr

Figure~\ref{fig:pe} shows the partial evaluator~|P|.
Its type |repr| literally expresses the type equation for |repr_pe| above.
The function |abstr|
extracts a future-stage code value from the result of
partial evaluation.  Conversely, the function |pdyn| injects a
code value into the |repr| type. As
in~\S\ref{S:PE-problem}, we build dynamic terms alongside
any static ones to avoid polymorphic lift.

The static portion of the interpretation of |lam f| is |Some f|, 
which just wraps the HOAS
function |f|. The interpretation of |app ef ea| 
checks to see if |ef| has such a wrapped
HOAS function. If it is, we apply |f| to the
concrete argument |ea|, giving us a chance to perform static
computations (see example the below). If
|ef| has only a dynamic part, we residualize.

\noindent All pattern\hyp
matching in~|P| is \emph{syntactically} exhaustive, so it is patent to the
metalanguage implementation that |P| never gets stuck.  Further, all
pattern\hyp matching occurs
during partial evaluation, only to check if a value is known statically,
never what type it has.  In other words, our partial evaluator tags
phases (with |Some| and |None|) but not object types.

%------------------------------------------------------------------------
% This is the longer version of the section
\else

\section{A tagless partial evaluator}\label{PE}
Surprisingly, we can write a partial evaluator using the idea above,
namely to build object terms using ordinary functions rather than data
constructors.  We present this partial evaluator in a sequence of four
attempts. It uses no universal type and no tags
for object types.  We then discuss residualization and binding-time
analysis.  Our partial evaluator is a modular extension of the evaluator
in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{compiler}, in
that it uses the former to reduce static terms and the latter to build
dynamic terms.

@@

Our typed partial evaluator is online and polyvariant.  It reuses the
compiler~|C| and the evaluator~|R| by composing them.  This situation is
simpler than \citets{SperberThiemann:TwoForOne} composition of a partial
evaluator and a compiler, but the general ideas are similar.
\fi
